Pattern Avoidance

### 2014-11-07 project2/paper/paper.tex
changed theorems in sections 3,4

\begin{corollary} 
For a permutation $\pi$, $s_n(\pi)=s_n(\mathcal{F}(\pi))$. 
\end{corollary}
\begin{proof}
This result follows because the flipping lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
\end{proof}
\section{Terms in the sequence $s_n(312)$}
\label{312}

@@

\begin{theorem}
$s_n(312)$, the total number of permutations of $\{1,2,3,...,n\}$ that avoid the pattern $312$, is equal to $C_n$, the $n^{th}$ Catalan number.
\end{theorem}

\begin{proof}
The inductive hypothesis holds for our base case of $\{1\}$, since the only permutation of $\{1\}$ trivially avoids $312$.

We need to prove the inductive case. Let us first assume that for all $i$ from $1$ to $k$, the number of permutations of $\{1,2,...,i\}$ that avoid the order $312$ as a subsequence is $C_i$.

@@

By the flipping lemma and reversing lemmas we see that  $s_n(132)$, $s_n(213)$ and $s_n(231)$ are the sequence of Catalan numbers as well.

@@

\section{Terms in the sequence $s_n(321)$}

\label{321}
Unfortunately, there is no clear way to reformulate Lemma \ref{lemma1} for the case of avoiding the permutation 321 or 123. As a consequence, the proof in the previous section simply does not hold. Fortunately, however, we still have a closed form expression for $s_n(321)$ (recall that $s_n(321)=s_n(123)$ by our flipping principle). Surprisingly, in fact, the sequence remains the same: $s_n(321)=C_n$, the $n^{th}$ Catalan number. In order to prove this result, we need the machinery of Young tableaux. What follows is a brief exposition of the key results, adapted from Stanley's \emph{Algebraic Combinatorics}.

@@

\begin{theorem}
\label{rsk}
There exists an algorithm (called the \emph{RSK algorithm}) which defines a bijection between the set of permutations in $S(n)$ avoiding a decreasing subsequence of length 3 and the set of pairs of size-$n$ SYT of the same shape and at most two rows. 
\end{theorem}\todo{give a citation for this}
Using all this machinery and all these results, we are finally able to prove the closed-form expression for $s_n(321)$.
\begin{theorem}
The total number of permutations of $\{1,2,3,\ldots, n\}$ that avoid the order 321 as a subsequence is $C_n$, where $C_n$ is the $n^{th}$ Catalan number.
\end{theorem}
\begin{proof}
By Theorem \ref{rsk}, there are $s_n(321)$ elements in the set $A$ of pairs of size-$n$ SYT of the same shape and at most two rows. By Theorem \ref{catalan_2}, the set $B_n$ of SYT of shape $(n, n)$ has its size given by $C_n$. We will construct a bijection between the sets $A_n$ and $B_n$: because $|A_n|=s_n(321)$ and $|B_n|=C_n$, it will therefore follow that $s_n(321)=C_n$, as desired.

>>>>

\begin{corollary} 
For a permutation $\pi$, $s_n(\pi)=s_n(\mathcal{F}(\pi))$. 
\end{corollary}
\begin{proof}
This result follows because the flipping lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
\end{proof}
\section{Avoiding Permutations of Length Three Part I}
\label{312}

@@

\begin{theorem}
$s_n(312), s_n(132), s_n(213)$, and $s_n(231)$ are equal to $C_n$, the $n^{th}$ Catalan number.
\end{theorem}

\begin{proof}
We prove this result for $(s_n(312))$ by induction and then use the flipping and reversing lemmas for the other three sequences. The inductive hypothesis holds for our base case of $\{1\}$, since the only permutation of $\{1\}$ trivially avoids $312$.

Let us assume that for all $i$ from $1$ to $k$, the number of permutations of $\{1,2,...,i\}$ that avoid the order $312$ as a subsequence is $C_i$.

@@

By the flipping lemma and reversing lemma we see that  $s_n(132)$, $s_n(213)$ and $s_n(231)$ are the sequence of Catalan numbers as well. 
\end{proof}

@@

\section{Avoiding Permutations of Length Three Part II}

\label{321}
Unfortunately, there is no clear way to reformulate Lemma \ref{lemma1} for the case of avoiding the permutation 321 or 123. As a consequence, the proof in the previous section simply does not hold. Fortunately, however, we still have a closed form expression for $s_n(321)$ (recall that $s_n(321)=s_n(123)$ by our flipping lemma). Surprisingly, in fact, the sequence remains the same: $s_n(321)=C_n$, the $n^{th}$ Catalan number. In order to prove this result, we need the machinery of Young tableaux. What follows is a brief exposition of the key results, adapted from Stanley's \emph{Algebraic Combinatorics}.

@@

\begin{theorem}
\label{rsk}
There exists an algorithm (called the \emph{RSK algorithm}) which defines a bijection between the set of permutations in $S(n)$ avoiding a decreasing subsequence of length 3 and the set of pairs of size-$n$ SYT of the same shape and at most two rows. 
\end{theorem}\todo{give a citation for this}
Using all this machinery and all these results, we are finally able to prove the closed-form expression for $s_n(321)$.
\begin{theorem}
$s_n(321)$ and $s_n(123)$ are equal to $C_n$, the $n^{th}$ Catalan number.
\end{theorem}
\begin{proof}
By the reversing lemma we only need to prove this result for $s_n(321)$. By Theorem \ref{rsk}, there are $s_n(321)$ elements in the set $A$ of pairs of size-$n$ SYT of the same shape and at most two rows. By Theorem \ref{catalan_2}, the set $B_n$ of SYT of shape $(n, n)$ has its size given by $C_n$. We will construct a bijection between the sets $A_n$ and $B_n$: because $|A_n|=s_n(321)$ and $|B_n|=C_n$, it will therefore follow that $s_n(321)=C_n$, as desired.
